---
title: "BIST5615 Homework 1"
subtitle: "Due date: Sunday November 3, 2024"
output: pdf_document
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-5em} 
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\vspace{-1cm}

## General Instructions

* Use this R Markdown template for homework submission.
* Answer the questions by inserting R code and necessary comments if applicable. Your output must contain the R code (do not use the `echo=FALSE` option) if applicable.
* Save the compiled PDF file under the file name `LastName-FirstName-HW1.pdf` 
and submit it through HuskyCT by the deadline. 

### Max Points: 35 points

### 1. For a $3\times 3$ table, please answer the following questions.


(a) (5 points) Assume that
\begin{align*}
P(X=1|Y=1)=0.5, & \;\; P(X=2|Y=1)=0.5, \;\; P(X=3|Y=1)=0, \\
P(X=1|Y=2)=0, & \;\; P(X=2|Y=2)=0.5, \;\; P(X=3|Y=2)=0.5, \\
P(X=1|Y=3)=0.5, & \;\; P(X=2|Y=2)=0, \;\; P(X=3|Y=3)=0.5; 
\end{align*}
and 
\begin{align*}
P(Y=1|X=1)=0.5, & \;\; P(Y=2|X=1)=0.5, \;\; P(Y=3|X=1)=0, \\
P(Y=1|X=2)=0, & \;\; P(Y=2|X=2)=0.5, \;\; P(Y=3|X=2)=0.5, \\
P(Y=1|X=3)=0.5, & \;\; P(Y=2|X=3)=0, \;\; P(Y=3|X=3)=0.5. 
\end{align*}
Can you uniquely determine the joint distribution of $X$ and $Y$?
Please provide the reason if not and find it if so.

(b) (5 points) Assume that
\begin{align*}
P(X=1|Y=1)=0.5, & \;\; P(X=2|Y=1)=0.5, \;\; P(X=3|Y=1)=0, \\
P(X=1|Y=2)=0, & \;\; P(X=2|Y=2)=0.5, \;\; P(X=3|Y=2)=0.5, \\
P(X=1|Y=3)=0.3, & \;\; P(X=2|Y=3)=0.3, \;\; P(X=3|Y=3)=0.4; 
\end{align*}
and 
\begin{align*}
P(Y=1|X=1)=0.5, & \;\; P(Y=2|X=1)=0.5, \;\; P(Y=3|X=1)=0, \\
P(Y=1|X=2)=0, & \;\; P(Y=2|X=2)=0.5, \;\; P(Y=3|X=2)=0.5, \\
P(Y=1|X=3)=0.5, & \;\; P(Y=2|X=3)=0, \;\; P(Y=3|X=3)=0.5. 
\end{align*}
Can you uniquely determine the joint distribution of $X$ and $Y$?
Please provide the reason if not and find it if so.

Ans.
Let's say \( X \) and \( Y \) take values in finite sets \( M \) and \( N \), and that their joint mass function \( \mathbb{P}[X=x, Y=y] \) is \( > 0 \) for all \( (x, y) \in M \times N \). Then:

\[
\mathbb{P}[X=x \mid Y=y] / \mathbb{P}[Y=y \mid X=x] = \mathbb{P}[X=x] / \mathbb{P}[Y=y],
\]

for all \( (x, y) \in M \times N \). Summing over \( x \), we find that:

\[
1 \cdot \mathbb{P}[Y=y] = \sum_{x \in M} \mathbb{P}[X=x \mid Y=y]/ \mathbb{P}[Y=y \mid X=x]
\]

is completely determined by the conditional distributions. Hence, so is \( \mathbb{P}[Y=y] \) determined. Finally, \( \mathbb{P}[X=x, Y=y] = \mathbb{P}[X=x \mid Y=y] / \mathbb{P}[Y=y] \) is also determined by the conditional distributions.

with this logic, for b)
\[
\begin{array}{|c|c|c|c|}
\hline
P(X = x, Y = y) & Y = 1 & Y = 2 & Y = 3 \\
\hline
X = 1 & 0 & 0 & \frac{3}{16} \\
X = 2 & 0 & \frac{3}{16} & \frac{3}{16} \\
X = 3 & 0 & \frac{3}{16} & \frac{4}{16} \\
\hline
\end{array}
\]



### 2. The sinking of the Titanic is a famous event, and new books are still being published about it.  Many well-known facts--from the proportions of first-class passengers to the ``women and children first" policy, and the fact that that policy was not entirely successful in saving the women and children in the third class--are reflected in the survival
rates for various classes of passenger.  These data were originally
collected by the British Board of Trade in their investigation of the
sinking.  The file titanicdat.csv contains the raw data (2201 observations, 4 variables). The file titanicdat.sas contains both the data and SAS codes. 
The file titanic.txt is a documentation file containing a brief description of the dataset. These files are posted on the Husky class website.

Data file: `titanic
```{r}
hw1q2 <-  read.csv("M:/teaching/BIST5615f24/HW/titanicdat.csv") 
str(hw1q2)
table(hw1q2)
```

(a) (5 points)  For adults who sailed on the Titanic, use the raw data to reproduce the odds of
survival for females was 11.4 times that of males as well as the odds of survival for females equaled 2.9 discussed in Lecture Note 3.

(b) (5 points) For adults who sailed on the Titanic, (i) compute the 95% exact confidence interval of the odds ratio of survival
for females versus males and (ii) compute a 95% asymptotic confidence interval and the 95% exact confidence interval of $\gamma$
for gender and survival status.

(c) (5 points) Using the entire raw data, determine whether the death rate for the third class passengers was much higher than for the others including the crew members.

(d) (5 points) Using the entire raw data, determine whether the death rate for females was much less than for males.

(e) (5 points) Using the entire raw data, determine whether the ``women and children first" policy was successful in
    saving the women and children.
