---
title: "cda_nutrition_project"
author: "Alokesh Manna"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

All the data is in XPT format. So need to have in a data frame. The explanation is in the doc file just after the data file. We are working with the data 2021-2023 (a very recent year.). There is another big dataset already available in R 2019-2014.

<https://rpubs.com/KamleshJha/Nhanes>

```{r}
library(NHANES)
library(dplyr)

```

```{r}
data(NHANES)
head(NHANES)
colnames(NHANES)
#tab <- NHANES %>% filter(AgeDecade == " 20-29", Gender == "female")
```


##lets work with the recnet one.
```{r}
# install.packages("haven")
# install.packages("writexl")
library(haven)
library(writexl)
library(dplyr)
library(tidyverse)
```

First we have the demography data downloaded.
```{r}
# Load the XPT file
#<https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Demographics&Cycle=2021-2023>
data <- read_xpt("~/downloads/DEMO_L.XPT")
df_demography=as.data.frame(data)
```


```{r}
head(df_demography)
```

```{r}
colnames(df_demography)
```







```{r}
summary(df_demography)
```



RIDAGEYR - Age in years at screening will be taken. Need to filter >17, ie 18 and above.

DMDEDUC2 - Education level - Adults 20+. Whether has a college education or not!

If it is 1,2,3 then no college education. If 4 and 5, then has college education.

RIDRETH1 - Race/Hispanic origin from 1 to 5, see from codebook
Value 4 is concerned. Non Hispanic black. 3	Non-Hispanic White
4	Non-Hispanic Black.

DMDYRUSR - Length of time in US. Immigration effects on blood pressure a lot. 

5006 people had college education. Non hispanic black are 1597. DMDYRUSR has 
10058 NA (did the just entered?)! 5575 male, 6358 female.
```{r}
df_demography|>
  filter(RIDAGEYR>17)|>
  mutate(DMDEDUC2=ifelse(DMDEDUC2>3,1,0))|> #pull(DMDEDUC2)|>table(), >3 college education
  mutate(RIDRETH1=ifelse(RIDRETH1==4,1,0))#|>#|>pull(RIDRETH1)|>table() ## Non Hispanic black are 1, rst 0
  #pull(RIAGENDR)|>table(), 1 male, 2 female, check from doc file, no NA
  
```




##definition in Demo_L_Doc

# SEQN - Respondent sequence number
# SDDSRVYR - Data release cycle
# RIDSTATR - Interview/Examination status
# RIAGENDR - Gender
# RIDAGEYR - Age in years at screening
# RIDAGEMN - Age in months at screening - 0 to 24 mos
# RIDRETH1 - Race/Hispanic origin from 1 to 5, see from codebook
# RIDRETH3 - Race/Hispanic origin w/ NH Asian
# RIDEXMON - Six-month time period
# RIDEXAGM - Age in months at exam - 0 to 19 years
# DMQMILIZ - Served active duty in US Armed Forces
# DMDBORN4 - Country of birth
# DMDYRUSR - Length of time in US
# DMDEDUC2 - Education level - Adults 20+
# DMDMARTZ - Marital status
# RIDEXPRG - Pregnancy status at exam
# DMDHHSIZ - Total number of people in the Household
# DMDHRGND - HH ref person’s gender
# DMDHRAGZ - HH ref person’s age in years
# DMDHREDZ - HH ref person’s education level
# DMDHRMAZ - HH ref person’s marital status
# DMDHSEDZ - HH ref person’s spouse’s education level
# WTINT2YR - Full sample 2-year interview weight
# WTMEC2YR - Full sample 2-year MEC exam weight
# SDMVSTRA - Masked variance pseudo-stratum
# SDMVPSU - Masked variance pseudo-PSU
# INDFMPIR - Ratio of family income to poverty
# National Health and Nutrition Examination Survey
# August 2021-August 2023 Data Documentation, Codebook, and Frequencies
# Demographic Variables and Sample Weights (DEMO_L)
# Data File: DEMO_L.xpt
# First Published: September 2024
# Last Revised: NA
# Component Description


Now blood pressure and cholestorol
```{r}
#bloodpressure and cholestorol
#<https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire&Cycle=2021-2023>
data_bloodpressure <- read_xpt("~/downloads/BPQ_L.XPT")
data_bloodpressure<-as.data.frame(data_bloodpressure)
colnames(data_bloodpressure)
```


BPQ020: {Have you/Has SP} ever been told by a doctor or other health professional that {you/s/he} had hypertension, also called high blood pressure?

BPQ030	{Were you/Was SP} told on 2 or more different visits that {you/s/he} had hypertension, also called high blood pressure?

BPQ080	{Have you/Has SP} ever been told by a doctor or other health professional that {your/his/her} blood cholesterol level was high?

BPQ150	{Are you/Is SP} now taking any medication prescribed by a doctor for {your/his/her/SP's} high blood pressure?

BPQ101D	{Are you/Is SP} now taking any medication prescribed by a doctor lower {your/his/her/SP's} blood cholesterol?

```{r}
head(data_bloodpressure)
```

```{r}
summary(data_bloodpressure)
```
```{r}
data_bloodpressure|>
  pull(BPQ020)|>
  table()
```


##final question

Q1: (more on blood pressure! fix one)
Does people have
association of race (black and white) and immigration status(only within black) 
with the development of chronic disease in USA? 

cholestorol (BPQ080),  ##almost no NA, only 3
blood pressure(BPQ020), ##almost no NA, only 3, 1 yes, 2 no, 7 refused, 10 donot know
diabatics (a complete other dataset, itself diabatics, diq010)


Just within black, can we compare between male/female?
categorize gender male, female # RIAGENDR - Gender


##solution of Q1

```{r}
df_modified_demography=
  df_demography|>
  mutate(clg_education=ifelse(DMDEDUC2>3,1,0))|> #pull(DMDEDUC2)|>table(), >3 college education
  mutate(clg_education=ifelse(clg_education=="NA",0,clg_education))|>##"NA" also no clg education
  mutate(nh_black=ifelse(RIDRETH1==4,1,0))|>#|>pull(RIDRETH1)|>table() ## Non Hispanic black are 1, rst 0
  mutate(nh_white=ifelse(RIDRETH1==3,2,0))|>## Non Hispanic white are 2, rst 0
  mutate(nh_black_white=nh_black+nh_white)
  #pull(RIAGENDR)|>table(), 1 male, 2 female, check from doc file, no NA
  
  
df_modified_bloodpressure=
  data_bloodpressure|>
  dplyr::select(c("SEQN","BPQ020"))|>
  mutate(blood_pressure=ifelse(BPQ020>1,2,1))|> ##if donot know or refused, lets say them no,##1 yes, 2 no.
 mutate(blood_pressure=ifelse(blood_pressure=="NA",2,blood_pressure))

df_modified_bloodpressure_demography=
  df_modified_bloodpressure|>
  left_join(df_modified_demography,by=c("SEQN"))


dim(df_modified_bloodpressure_demography)
```

```{r}
colnames(df_modified_bloodpressure_demography)
```


```{r}
df_modified_bloodpressure|>pull(blood_pressure)|>table()
```


```{r}
df_modified_bloodpressure_demography|>
  dplyr::select(c("nh_black_white","SEQN","blood_pressure","clg_education","RIAGENDR","DMDYRUSR","RIDAGEYR","DMDMARTZ"))|>
  #na.omit()|> ## still some NA, so just drop them
  summary()
```
```{r}
##nh black=1, white=2, other=0
##bp=1 yes, 2 no
##clg_education= 1 yes, 0 no
##gender RIAGENDR=1 male 2 female 
#DMDMARTZ - Marital status 1	Married/Living with partner	,2	Widowed/Divorced/Separated
#3	Never married,77	Refused,99	Don't know
nh_bp_edu_gender_immigration=
  df_modified_bloodpressure_demography|>
  dplyr::select(c("nh_black_white","SEQN","blood_pressure","clg_education","RIAGENDR","DMDYRUSR","RIDAGEYR","DMDMARTZ"))|>
  mutate(
    nh_black_white=as.factor(nh_black_white),
    blood_pressure=as.factor(blood_pressure),
    clg_education=as.factor(clg_education),
    RIAGENDR=as.factor(RIAGENDR),
    DMDMARTZ=as.factor(DMDMARTZ)
  )|>
  
  na.omit()
```


```{r}
nh_bp_edu_gender_immigration
```

Can we do some regression techniques? nh_black_white has also 0.
```{r}
model_reg1 <- glm(
  blood_pressure ~ (nh_black_white 
  +clg_education+ RIAGENDR+ DMDYRUSR
  +RIDAGEYR
  )
    
    , family = binomial(link = "logit"), 
  
  data = nh_bp_edu_gender_immigration
  )

```

```{r}
summary(model_reg1)
```




```{r}
model_reg2 <- glm(
  blood_pressure ~ (nh_black_white 
  +clg_education+ RIAGENDR+ DMDYRUSR
  +RIDAGEYR
  )^2
    
    , family = binomial(link = "logit"), 
  
  data = nh_bp_edu_gender_immigration|>
    filter(nh_black_white!=0)
  )

```

```{r}
summary(model_reg2)
```

```{r}
anova(model_reg2,test = "Chisq")
```

```{r}
library(pROC)
actual_outcomes<-
  
  nh_bp_edu_gender_immigration|>
    filter(nh_black_white!=0)|>
  pull(blood_pressure)

fitted_values<-
  
  model_reg2$fitted.values
  
  
roc_curve <- roc(actual_outcomes, fitted_values)
auc(roc_curve)
plot(roc_curve, main = "ROC Curve for only non hispanic population")

```
```{r}
library(pROC)
actual_outcomes<-
  
  nh_bp_edu_gender_immigration|>
    #filter(nh_black_white!=0)|>
  pull(blood_pressure)

fitted_values<-
  
  model_reg1$fitted.values
  
  
roc_curve <- roc(actual_outcomes, fitted_values)
auc(roc_curve)
plot(roc_curve, main = "ROC Curve for overall population")

```


```{r}
library(logistf)

model_f=logistf(blood_pressure ~ nh_black_white + clg_education
  +clg_education+RIAGENDR+DMDYRUSR
    
    , family = binomial(link = "logit"), 
  
  data = nh_bp_edu_gender_immigration)

summary(model_f)
```




```{r}
nh_bp_gender_table <- table(
  nh_bp_edu_gender_immigration[,"blood_pressure"], 
  nh_bp_edu_gender_immigration[,"RIAGENDR"]
  )
#table(nh_bp_gender[,"blood_pressure"])
nh_bp_gender_table
addmargins(nh_bp_gender_table)
```

```{r}
nh_bp_race_table <- table(nh_bp_edu_gender_immigration[,"blood_pressure"], nh_bp_edu_gender_immigration[,"nh_black_white"])

addmargins(nh_bp_race_table)
```
what if we filter nh_black_white>0? just NH black and white.
```{r}
nh_only_black_white=nh_bp_edu_gender_immigration|>
  filter(nh_black_white!=0)|>
  mutate(nh_black_white=factor(nh_black_white))

nh_bp_race_B_or_W_table <- table(nh_only_black_white[,"blood_pressure"], nh_only_black_white[,"nh_black_white"])

addmargins(nh_bp_race_B_or_W_table)
```


```{r}
nh_bp_immigration_table <- table(nh_bp_edu_gender_immigration[,"blood_pressure"], nh_bp_edu_gender_immigration[,"DMDYRUSR"])

addmargins(nh_bp_immigration_table)
```


a) odds ratio test: fisher

```{r}

fisher_test <- fisher.test(nh_bp_gender_table)
print(fisher_test$estimate)  # Odds ratio
print(fisher_test$conf.int)  # Confidence interval for the odds ratio
fisher_test
```
only in non hispanic the p value is small! So within non hispanic, there might be association between bp and black and white.
```{r}

fisher_test <- fisher.test(nh_bp_race_B_or_W_table)
print(fisher_test$estimate)  # Odds ratio
print(fisher_test$conf.int)  # Confidence interval for the odds ratio
fisher_test
```



b) epitools

```{r}
library(epitools)

# Calculate odds ratio with confidence interval
odds_ratio_result <- oddsratio(nh_bp_gender_table)
print(odds_ratio_result)
```






c) gamma:1 indicates a perfect positive association, -1 indicates a perfect negative association, 0 suggests no association.

```{r}
library(vcdExtra)
gamma_result <- GKgamma(nh_bp_gender_table)
print(gamma_result)
```

race and bp are not related.
```{r}
library(vcdExtra)
gamma_result <- GKgamma(nh_bp_race_table)
print(gamma_result)
```

only in non-hispanic
```{r}

library(vcdExtra)
gamma_result <- GKgamma(nh_bp_race_B_or_W_table)
print(gamma_result)
```




```{r}
library(MESS)
gkgamma(nh_bp_race_table, conf.level=0.95)
```


bp and immigration are related.
```{r}

library(vcdExtra)
gamma_result <- GKgamma(nh_bp_immigration_table)
print(gamma_result)
```

```{r}
library(MESS)
gkgamma(nh_bp_immigration_table, conf.level=0.95)
```



##1.categrize Age yong adults(18-39), middle age (40-59), older adults (>60)
(Not now!)
##2. categorize gender male, female # RIAGENDR - Gender
##3. having high blood pressure greater than 2 times in doctor visit. yes no
###in black. origin category, what is the prevalence of 
#high bloodpressure in young adults vs older adults

##4. may compare black and white

##<https://wwwn.cdc.gov/Nchs/Nhanes/2021-2022/DBQ_L.htm> diet nutrition data

#go here <https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire> and search for DBQ_L
```{r}
data <- read_xpt("~/downloads/DBQ_L.XPT")
df_diet=as.data.frame(data)
```

```{r}
head(df_diet); dim(df_diet)
```

```{r}
df_self_meal_prepare=
df_diet|>
  dplyr::select(c("SEQN","DBQ930"))|> ##1 yes, 2 no, 9 donot know
  filter(DBQ930!=9)|>
  mutate(self_meal_prepare=as.factor(DBQ930))#|>pull(self_meal_prepare)|>table()
  

df_self_meal_prepare|>pull(self_meal_prepare)|>table()
```

```{r}
df_bp_self_meal_prepare=
df_modified_bloodpressure|>
  left_join(df_self_meal_prepare,by=c("SEQN"))

table(df_bp_self_meal_prepare$blood_pressure,df_bp_self_meal_prepare$self_meal_prepare)|>
  addmargins()

df_bp_self_meal_prepare_table=

table(df_bp_self_meal_prepare$blood_pressure,
      df_bp_self_meal_prepare$self_meal_prepare)
```

```{r}
fisher_test <- fisher.test(df_bp_self_meal_prepare_table)
```

```{r}
fisher_test
```
```{r}
library(MESS)
gkgamma(df_bp_self_meal_prepare_table, conf.level=0.95)
```


```{r}
df_bp_self_meal_prepare
```


## for Q2
```{r}
dat=na.omit(df_bp_self_meal_prepare)

mod_bp_meal=glm(as.factor(blood_pressure)~self_meal_prepare,
     family = binomial(link = "logit"), 
  
  data = dat
    
)
```


```{r}
summary(mod_bp_meal)
```


```{r}
library(pROC)
actual_outcomes<- dat$blood_pressure

fitted_values<- mod_bp_meal$fitted.values
  
  
roc_curve <- roc(actual_outcomes, fitted_values)

#auc(roc_curve)
plot(roc_curve)

```
whats about to include all variables?

```{r}
#nh_bp_edu_gender_immigration

df_nh_bp_edu_gender_immigration_self_meal_prepare=
nh_bp_edu_gender_immigration|>
  left_join(df_self_meal_prepare,by=c("SEQN"))

df_nh_bp_edu_gender_immigration_self_meal_prepare
```
```{r}
colnames(df_nh_bp_edu_gender_immigration_self_meal_prepare)
```


```{r}
mod_nh_bp_meal_1int<- 
  glm( as.factor(blood_pressure)~(nh_black_white 
  +clg_education + RIAGENDR+ DMDYRUSR+ RIDAGEYR + self_meal_prepare ),
     family = binomial(link = "logit"), 
  
  data = na.omit(df_nh_bp_edu_gender_immigration_self_meal_prepare)
#|>filter(nh_black_white!=0) 
)
```



```{r}
summary(mod_nh_bp_meal_1int)
```

```{r}
mod_nh_bp_meal_2int<- 
  glm( as.factor(blood_pressure)~(nh_black_white 
  +clg_education + RIAGENDR+ DMDYRUSR+ RIDAGEYR + self_meal_prepare )^2,
     family = binomial(link = "logit"), 
  
  data = na.omit(df_nh_bp_edu_gender_immigration_self_meal_prepare)
#|>filter(nh_black_white!=0) 
)

summary(mod_nh_bp_meal_2int)
```




Q2. Blood pressure 0/1 (BPQ020). 
DBQ930 - Main meal planner/preparer: {Are you/Is SP} the person who does 
most of the planning or preparing of meals in {your/SP's} family?0/1
If you do not prepare meal by yourself, you can have hypertension as somebody else who is preparing food for you, might not know about your health condition. You can also eat junk food!

Soln: Based on the above analysis bp and meal preparation are associated.

##problem 2:
#DBQ930 - Main meal planner/preparer
##if black people do prepare food at home or outside while having hypertention
##link age, gender, meal, education level


Q3: Maybe this variables are a bit complicated, we try other data
Take 2 variables:
PAD790Q - Frequency of moderate LTPA,
PAD810Q - Frequency of vigorous LTPA

If it is greater than 150 min per week of moderate PA or 75 min per week of vigorous PA

If black adult does 150 minuites or more Physical Activity (PA), then prevalence
of hypertension. 


https://wwwn.cdc.gov/Nchs/Nhanes/2021-2022/PAQ_L.htm
Find the data in questioaire


##problem 3:
#https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire for physical activity , search PAQ_L.XPT
#PAD800 - Minutes moderate LTPA how much to do exercisein leisure time
##check if black adults had enough physical activity and blood pressure relation
#

```{r}
data_physical_activity <- read_xpt("~/downloads/PAQ_L.XPT")
data_physical_activity<-as.data.frame(data_physical_activity)
colnames(data_physical_activity)
```


```{r}
df_modified_physical_activity=
data_physical_activity|>
  mutate(moderate_LTPA=PAD790Q,
         vigorous_LTPA=PAD810Q )|>
  mutate(total_activity=moderate_LTPA+vigorous_LTPA)|>
  mutate(total_activity_index=ifelse(total_activity>150,1,0))|>#|>summary()
  dplyr::select(c("SEQN","total_activity_index"))|>
  mutate(total_activity_index=as.factor(total_activity_index))

df_modified_physical_activity
  
```


```{r}
df_bp_activity=
df_nh_bp_edu_gender_immigration_self_meal_prepare|>
#df_modified_bloodpressure|>
  left_join(df_modified_physical_activity,by=c("SEQN"))|>
# |>
   dplyr::select(c("SEQN","total_activity_index","blood_pressure", "nh_black_white", "clg_education",
  "clg_education", "RIAGENDR", "DMDYRUSR", "DMDMARTZ",
  "RIDAGEYR"))

table(df_bp_activity$blood_pressure,df_bp_activity$total_activity_index)|>
  addmargins()

df_bp_activity_table=

table(df_bp_activity$blood_pressure,df_bp_activity$total_activity_index)
```

```{r}
library(MESS)
gkgamma(df_bp_activity_table, conf.level=0.95)
```

```{r}
fisher_test <- fisher.test(df_bp_activity_table)
fisher_test
```

so in general, no association. Can we check for black NH?



```{r}
df_bp_activity
```

```{r}
df_bp_activity_modified=
  na.omit(df_bp_activity)|>
  mutate(blood_pressure=as.factor(blood_pressure))
  

mod_activity=
  glm(
  blood_pressure~total_activity_index,
    
  family = binomial(link = "logit"), 
  
  data =df_bp_activity_modified
  )
```

```{r}
summary(mod_activity)
```


```{r}
library(pROC)
actual_outcomes<-
  
df_bp_activity_modified$blood_pressure

fitted_values<-
  
  mod_activity$fitted.values
  
  
roc_curve <- roc(actual_outcomes, fitted_values)
#auc(roc_curve)
plot(roc_curve)

```

## within the black population
```{r}
mod_activity_withOtherCovariates_1int=
  glm(
  as.factor(blood_pressure) ~ (nh_black_white 
  +clg_education+ RIAGENDR+ DMDYRUSR
  +RIDAGEYR+total_activity_index
  ),
    
  family = binomial(link = "logit"), 
  
  data =df_bp_activity_modified #|>filter(nh_black_white==1)
  )

mod_activity_withOtherCovariates_2int=
  glm(
  as.factor(blood_pressure) ~ (
    nh_black_white 
  +clg_education+ RIAGENDR+ DMDYRUSR
  +RIDAGEYR+total_activity_index
  )^2,
    
  family = binomial(link = "logit"), 
  
  data =df_bp_activity_modified #|>filter(nh_black_white==1)
  )
```


```{r}
summary(mod_activity_withOtherCovariates_1int)
```



```{r}
summary(mod_activity_withOtherCovariates_2int)
```


Can we use lightgbm? some tree based approach? 
```{r}
colnames(df_bp_activity_modified)
dim(df_bp_activity_modified)
set.seed(42)
train_idx <- sample(1:nrow(df_bp_activity_modified), size = 0.8 * nrow(df_bp_activity_modified))
train_data <-df_bp_activity_modified[train_idx, ]
test_data <- df_bp_activity_modified[-train_idx, ]

# Separate features and target variable
train_X <- as.matrix(train_data[, -c(1,3)])
train_y <- as.numeric(train_data[, 3])-1

test_X <- as.matrix(test_data[, -c(1,3)])
test_y <- as.numeric(test_data[, 3])-1#
colnames(test_X)
```

```{r}
library(lightgbm)

# Create LightGBM Dataset for training and testing
train_lgb <- lgb.Dataset(data = train_X, label = train_y, categorical_feature = c(1:4,6))
test_lgb <- lgb.Dataset(data = test_X, label = test_y)

```



```{r}
params <- list(
  objective = "multiclass", 
  num_class = 3, 
  metric = "multi_logloss", 
  boosting_type = "gbdt", 
  num_leaves = 31, 
  learning_rate = 0.05, 
  feature_fraction = 0.9, 
  bagging_fraction = 0.8, 
  bagging_freq = 5
)

```



```{r}
# Train the model
model <- lgb.train(
  params = params,
  data = train_lgb,
  nrounds = 100,                # 'num_round' is now 'nrounds'
  early_stopping_rounds = 10,    # Early stopping works as before
  valids = list(test = test_lgb) # 'valid_sets' is now 'valids'
)


```
```{r}
# Make predictions
pred <- predict(model, test_X)
pred_labels <- max.col(pred) - 1  # Convert to class labels

# Evaluate accuracy
accuracy <- sum(pred_labels == test_y) / length(test_y)
print(paste("Accuracy:", round(accuracy, 4)))
```

```{r}
# Get feature importance
importance <- lgb.importance(model)

# View the feature importance
print(importance)
lgb.plot.importance(importance, top_n = 20, measure = "Gain")

```






```{r}
# Make predictions (probabilities) on the test set
pred_probs <- predict(model, test_X)

# View the predicted probabilities
head(pred_probs)

# Load pROC for AUC calculation
library(pROC)

# Convert predictions to factors
pred_class <- max.col(pred_probs) - 1  # The class with the highest probability

# AUC for each class
roc_curve_class0 <- roc(test_y == 0, pred_probs[, 1])  # Class 0 (Setosa)
roc_curve_class1 <- roc(test_y == 1, pred_probs[, 2])  # Class 1 (Versicolor)
#roc_curve_class2 <- roc(test_y == 2, pred_probs[, 3])  # Class 2 (Virginica)

# AUC values
cat("AUC for Class 0 (Setosa): ", auc(roc_curve_class0), "\n")
cat("AUC for Class 1 (Versicolor): ", auc(roc_curve_class1), "\n")
#cat("AUC for Class 2 (Virginica): ", auc(roc_curve_class2), "\n")

```
```{r}
plot(roc_curve_class0)
plot(roc_curve_class1)
```



##Can we add a few more data?
a) Income data
b) Marital status


```{r}

```



##Some neural net?
```{r}
# install.packages("keras3")
# keras3::install_keras(backend = "tensorflow")
```




```{r}
library(keras3)

```

```{r}
# One-hot encode categorical variables (e.g., nh_black_white, clg_education)
df_bp_activity_modified <- df_bp_activity_modified %>%
  mutate_at(vars(nh_black_white, clg_education), as.factor) %>%
  mutate_at(vars(nh_black_white, clg_education), function(x) as.numeric(factor(x)))

# Normalize continuous variables (e.g., RIDAGEYR, total_activity_index, etc.)
continuous_vars <- c("RIDAGEYR",  "DMDYRUSR")
df_bp_activity_modified[continuous_vars] <- scale(df_bp_activity_modified[continuous_vars])

```


```{r}
# Set the target variable
y <- df_bp_activity_modified$blood_pressure

# Set the predictor variables (excluding the target)
X <- df_bp_activity_modified %>%
  dplyr::select(c(nh_black_white, clg_education, RIAGENDR, DMDYRUSR, RIDAGEYR, total_activity_index, DMDMARTZ))

# Convert to matrix for neural network
# X <- as.matrix(X)
# y <- as.factor(y)
head(X)
```

```{r}
X=X|>
  mutate(RIAGENDR=as.numeric(RIAGENDR),
         total_activity_index=as.numeric(total_activity_index),
         DMDMARTZ=as.numeric(DMDMARTZ))

X=as.matrix(X)
head(X)
```


```{r}
# Define the model
model <- keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = ncol(X)) %>%  # First hidden layer
  layer_dropout(rate = 0.2) |>
  layer_dense(units = 32, activation = "relu") %>%                        # Second hidden layer
    layer_dropout(rate = 0.2) |>
  layer_dense(units = 1, activation = "sigmoid")                          # Output layer (binary classification)

# Compile the model
model %>% compile(
  loss = "binary_crossentropy",      # For binary classification
  optimizer = "adam",                # Adam optimizer
  metrics = c("accuracy")            # Track accuracy during training
)

```



```{r}
history <- model %>% fit(
  X, y,
  epochs = 60,           # Number of training epochs
  batch_size = 20,       # Batch size
  validation_split = 0.2, # 20% of data used for validation
  verbose = 1            # Display progress during training
)

```

```{r}
# Predict probabilities
predicted_probs <- model %>% predict(X)

# Convert probabilities to binary outcomes (use a threshold of 0.5)
predicted_classes <- ifelse(predicted_probs > 0.5, 2, 1)

# Evaluate the accuracy
accuracy <- mean(predicted_classes == y)
print(paste("Accuracy: ", accuracy))

```


```{r}
library(pROC)
actual_outcomes<-
  
df_bp_activity_modified$blood_pressure

fitted_values<-predicted_probs
  
  
roc_curve <- roc(actual_outcomes, fitted_values)
auc(roc_curve)
plot(roc_curve)

```


```{r}
# Extract weights of the first layer
weights <- model %>% get_weights()

# Weights of the first layer (weights and biases)
first_layer_weights <- weights[[1]]  # First layer weights matrix
first_layer_biases <- weights[[2]]   # Biases of the first layer

# Visualize the weight distribution using a histogram
ggplot(data.frame(weights = as.vector(first_layer_weights)), aes(x = weights)) +
  geom_histogram(binwidth = 0.05, fill = 'blue', color = 'black', alpha = 0.7) +
  labs(title = "Weight Distribution of First Layer", x = "Weight Values", y = "Frequency") +
  theme_minimal()

```




```{r}
# Visualize the weight distribution using a histogram
ggplot(data.frame(biases = as.vector(first_layer_biases)), aes(x = biases)) +
  geom_histogram(binwidth = 0.05, fill = 'blue', color = 'black', alpha = 0.7) +
  labs(title = "Weight Distribution of First Layer", x = "Weight Values", y = "Frequency") +
  theme_minimal()
```

